---
output:
  github_document
---

```{r echo=FALSE}
set.seed(42)
options(digits=2)
```
<!-- README.md is generated from README.Rmd. Please edit that file -->

# DeclareDesign: Declare and diagnose research designs to understand and improve them.

**DeclareDesign** is statistical software that makes it easier for researchers to characterize and learn about the properties of research designs before implementation. Ex ante declaration and diagnosis of designs can help researchers clarify the strengths and limitations of their designs and to improve their properties. It can make it easier for readers to evaluate a research strategy prior to implementation and without access to results. It can also make it easier for designs to be shared, replicated, and critiqued.

The motivation for the software is described in a [working paper](http://declaredesign.org/paper.pdf) by the authors.

**DeclareDesign** consists of a core package, which is documented on this web site, as well as three companion packages that stand on their own but are also called on by the core package. They are:

1. [randomizr](http://randomizr.declaredesign.org): Easy to use tools for common forms of random assignment and sampling.
2. [fabricatr](http://fabricatr.declaredesign.org): Imagine your data before you collect it.
3. [estimatr](http://estimatr.declaredesign.org): Fast estimators for social scientists.

A [library](http://declaredesign.org/articles/design_library.html) of declared designs is under construction. The library includes canonical designs which users can download, modify, and deploy.

**Note:** *This software is not yet ready for general release. Please contact the authors before using in experiments or published work. Specifications, names, and arguments of functions are subject to change.*

## Installing DeclareDesign

To install the latest development release of all of the packages, please ensure that you are running version 3.3 or later of R and run the following code:

```{r, eval=F}
install.packages("DeclareDesign", dependencies = TRUE, 
                 repos = c("http://R.declaredesign.org", "https://cloud.r-project.org"))
```

## Getting started

Once **DeclareDesign** is installed, it is easy to begin working with your own research designs. The below example implements a simple two-arm (treatment and control) experiment on a population. We simulate population level data and ensure that the treatment effect should be "1 unit" on average. We then sample a sample from this population, randomly assign members of our sample to treatment, and calculate an average treatment effect on the resulting sample. We use Monte Carlo simulation to decide whether, under repeated experiments, we are able to detect the effect we assumed in our population, whether our estimates have bias, what our power to detect a "statistically significant" effect is, and what statistical coverage our estimator displays:

```{r echo=TRUE, results="hide", eval=FALSE}
library(DeclareDesign)

# Model a population with 500 units
population <- declare_population(N = 500, 
                                 noise = rnorm(N), 
                                 treatment_effect = rnorm(N, mean=1))
potential_outcomes <- declare_potential_outcomes(Y_Z_0 = noise, 
                                                 Y_Z_1 = noise + treatment_effect)

estimand <- declare_estimand(ATE = mean(Y_Z_1 - Y_Z_0))

# Data Strategy
sampling <- declare_sampling(n = 250) # Sample of 250 units
assignment <- declare_assignment(m = 100) # 100 treated units

# Answer Strategy
estimator <- declare_estimator(Y ~ Z, estimand = estimand)

# Design
two_arm <- declare_design(population, 
                          potential_outcomes, 
                          sampling, 
                          estimand, 
                          assignment, 
                          reveal_outcomes, 
                          estimator)

# Diagnose
diagnosis <- diagnose_design(two_arm, 
                             sims = 1000, 
                             bootstrap_sims = 500)

diagnosis
```
```{r echo=FALSE}
# To speed index building
temp_d = c("ATE", "my_estimator", -0.002751735, 
           0.004804101, 0.1501357, 0.003382338, 
           1, 0, 0.967,
           0.005855739, 0.9971695, 0.005274545, 
           0.1646188, 0.003860809, 0, 
           0, 0.9999212, 0.002032432)
names(temp_d) = c("Estimand Label", "Estimator Label", "Bias", 
                  "SE(Bias)", "RMSE", "SE(RMSE)", 
                  "Power", "SE(Power)", "Coverage", 
                  "SE(Coverage)", "Mean Estimatr", "SE(Mean Estimate)",
                  "SD Estimate", "SE (SD Estimate)", "Type S Rate", 
                  "SE(Type S Rate)", "Mean Estimand", "SE(Mean Estimand)")

result = t(data.frame(temp_d))
rownames(result) = NULL
knitr::kable(result)
```

This project is generously supported by a grant from the [Laura and John Arnold Foundation](http://www.arnoldfoundation.org) and seed funding from [EGAP](http://egap.org).
