---
output:
  github_document
title: "DeclareDesign: Declare and diagnose research designs to understand and improve them"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE}
set.seed(42)
options(digits=2)
```

[![Travis-CI Build Status](https://travis-ci.org/DeclareDesign/DeclareDesign.svg?branch=master)](https://travis-ci.org/DeclareDesign/DeclareDesign)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/DeclareDesign/DeclareDesign?branch=master&svg=true)](https://ci.appveyor.com/project/DeclareDesign/DeclareDesign)
[![Coverage Status](https://coveralls.io/repos/github/DeclareDesign/DeclareDesign/badge.svg?branch=master)](https://coveralls.io/github/DeclareDesign/DeclareDesign?branch=master)


## Overview
DeclareDesign is a system for describing research designs in code and simulating them in order to understand their properties. Because DeclareDesign relies a consistent grammar of designs, you can focus on the intellectually challenging part -- designing good research studies -- without having to code up simulations from scratch.

## Installation

To install the latest development release of all of the packages, please ensure that you are running version 3.4 or later of R and run the following code:

```{r, eval=F}
install.packages("DeclareDesign", dependencies = TRUE,
                 repos = c("http://R.declaredesign.org", "https://cloud.r-project.org"))
```

## Usage

Designs are declared by adding together design elements. Here's a minimal example that describes a 100 unit randomized controlled trial with a binary outcome. Half the units are assigned to treatment and the remainder to control. The true value of the average treatment effect is 0.05 and it will be estimated with the difference-in-means estimator. The diagnosis shows that the study is unbiased but underpowered.

```{r, eval=TRUE, message=FALSE}
library(DeclareDesign)

design <-
  declare_population(N = 100) +
  declare_potential_outcomes(Y ~ rbinom(n = N, size = 1, prob = 0.5 + 0.05 * Z)) +
  declare_estimand(ATE = 0.05) +
  declare_assignment(m = 50) +
  declare_estimator(Y ~ Z)

diagnosis <- diagnose_design(design, diagnosands = declare_diagnosands(select = c("power", "bias")))
diagnosis
```

## Philosophy

You can read a short post on the the idea behind DeclareDesign [here](https://declaredesign.org/idea/). A fuller description of the philosophy underlying the software is described in this [working paper](https://declaredesign.org/paper.pdf).

## Companion software

The core DeclareDesign package relies on three companion packages, each of which is useful in its own right.

1. [randomizr](https://declaredesign.org/R/randomizr/): Easy to use tools for common forms of random assignment and sampling.
2. [fabricatr](https://declaredesign.org/R/fabricatr/): Imagine your data before you collect it.
3. [estimatr](https://declaredesign.org/R/estimatr/): Fast estimators for social scientists.

## Learning DeclareDesigns

To get underway with your own designs, check out our guide on [getting started with DeclareDesign](https://declaredesign.org/R/DeclareDesign/articles/DeclareDesign.html), which covers the main functionality of the software.

You can also browse a [library](https://declaredesign.org/library/) of already declared designs. The library includes canonical designs that you can download, modify, and deploy.

---

This project is generously supported by a grant from the [Laura and John Arnold Foundation](http://www.arnoldfoundation.org) and seed funding from [EGAP](http://egap.org).
