---
title: "Custom functions and DeclareDesign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom functions and DeclareDesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
set.seed(42)
library(DeclareDesign)
options(digits=2)

my_population <-
  declare_population(N = 1000,
  income = rnorm(N),
  age = sample(18:95, N, replace = TRUE))

pop <- my_population()

my_potential_outcomes <- declare_potential_outcomes(
  formula = Y ~ .25 * Z + .01 * age * Z)
pop_pos <- my_potential_outcomes(pop)

my_sampling <- declare_sampling(n = 250)
smp <- my_sampling(pop_pos)

my_assignment <- declare_assignment(m = 25)
smp <- my_assignment(smp)

my_estimand <- declare_estimand(ATE = mean(Y_Z_1 - Y_Z_0))

smp <- reveal_outcomes(smp)
```

Because all inputs to `declare_design()` are functions, it's easy for you to provide custom functions instead of using our defaults.

# Custom Population

You can use a custom function to generate your population entirely on your own, too:

```{r echo=TRUE, results="hide"}
my_population_function <- function(N) {
  data.frame(u = rnorm(N))
}

my_population_custom <- declare_population(
  handler = my_population_function, N = 100)

pop_custom <- my_population_custom()

head(pop_custom)
```
```{r echo=FALSE}
knitr::kable(head(pop_custom))
```

# Custom Potential Outcomes

```{r echo=TRUE, results="hide"}
my_potential_outcomes_function <-
  function(data) {
    data$Y_Z_0 <- with(data, u)
    data$Y_Z_1 <- with(data, 0.25 + u)
    data
  }
my_potential_outcomes_custom <- declare_potential_outcomes(
  handler = my_potential_outcomes_function
)

pop_pos_custom <- my_potential_outcomes_custom(pop_custom)

head(pop_pos_custom[, c("u", "Y_Z_0", "Y_Z_1")])
```
```{r echo=FALSE}
knitr::kable(head(pop_pos_custom[, c("u", "Y_Z_0", "Y_Z_1")]))
```

# Custom Sampling

Again, you can still use a custom sampling function easily. In this case, the requirement is simply that the function  takes population data and returns sampled data. You can also include inclusion weights if you wish in the function (as the default function does). 

```{r echo=TRUE, results="hide"}
my_sampling_function <- function(data) {
     data$S <- rbinom(n = nrow(data),
            size = 1,
            prob = 0.1)
     data[data$S == 1, ]
}

my_sampling_custom <- declare_sampling(
  handler = my_sampling_function)

smp_custom <- my_sampling_custom(pop_pos)

nrow(smp_custom)
```
`r nrow(smp_custom)`

# Custom Assignment

```{r echo=TRUE, results="hide"}
my_assignment_function <- function(data) {
  data$Z <- rbinom(n = nrow(data),
         size = 1,
         prob = 0.5)
  data
}
my_assignment_custom <- declare_assignment(
  handler = my_assignment_function)

table(my_assignment_custom(pop_pos)$Z)
```
```{r echo=FALSE}
knitr::kable(t(as.matrix(table(my_assignment_custom(pop_pos)$Z))))
```

# Custom Estimand

Handlers can also make use of labels, if it's declared as an argument to the custom handler function.

```{r echo=TRUE, results="hide"}
my_estimand_function <- function(data, label) {
    data.frame(
      estimand_label=label,
      estimand=with(data, median(Y_Z_1 - Y_Z_0))
    )
}
my_estimand_custom <- declare_estimand(
  handler = my_estimand_function, label = "medianTE")

my_estimand_custom(pop_pos)
```
```{r echo=FALSE}
knitr::kable(my_estimand_custom(pop_pos))
```

# Custom Estimator

Custom estimators are slightly different - the default handler itself calls a model function. To implement a new handler, you can use the tidy_estimator
wrapper to handle the bookkeeping associated with estimator and estimand labeling.

```{r echo=TRUE, results="hide"}
my_estimator_function <- function(data){
  data.frame(est = with(data, mean(Y)))
}

my_estimator_custom <- 
  declare_estimator(handler = tidy_estimator(my_estimator_function), 
                    estimand = my_estimand)

my_estimator_custom(smp)
```
```{r echo=FALSE}
knitr::kable(my_estimator_custom(smp))
```
