---
title: "DeclareDesign 2.0: what changes, what does not, and why"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DeclareDesign 2.0: what changes, what does not, and why}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      purl = FALSE)
```

## The short version

DeclareDesign 2.0 is a ground-up reimplementation of DeclareDesign on tidyverse primitives. The public API is the one you already know: `declare_model()`, `declare_inquiry()`, `declare_sampling()`, `declare_assignment()`, `declare_measurement()`, `declare_estimator()`, joined by `+`, then handed to `simulate_design()`, `diagnose_design()` and `redesign()`.

Three things are true of the rewrite, and this vignette defends each in turn.

Most designs run unchanged. Of the 90 design declarations in the DeclareDesign book, 81 are checked in as tests and every one of them builds, runs and returns estimates; the accounting for the other nine is in [Current status] below. What a typical declaration has to change is one thing: an estimator's `inquiry =` is the inquiry's label, not the step object. Everything else that changes is at the edges (a handful of argument names, three functions that went, one way of grouping a diagnosis), and every one of those changes stops with an error naming the replacement rather than running to a different answer.

The pipeline is faster and it parallelises. Simulation and diagnosis run 1.15 to 1.35 times faster sequentially, and four workers make a 2,000-simulation run 3.2 times faster after one line of `future::plan()`. A seeded run gives the same table sequentially and in parallel.

The rewrite can express designs the original cannot. A design carries the objects it reads, so it survives `saveRDS()` and travels to a worker. `declare_parameters()` names what a design can be redesigned over, and `declare_notes()` names what it works out along the way. Step-level `draws` hold a population fixed while the assignment is redrawn, or the reverse, and the diagnosis then decomposes the variance of any per-simulation quantity by the step that generated it. An estimator that fails on one draw is recorded rather than fatal.

## Getting it

The 2.0 packages are not on CRAN yet. They install from GitHub under their released names, so installing them replaces the CRAN versions in your library; keep the CRAN versions available by installing them into a separate library and passing `lib.loc` when you want them.

```{r}
remotes::install_github("DeclareDesign/fabricatr@rewrite")
remotes::install_github("DeclareDesign/estimatr@rewrite")
remotes::install_github("DeclareDesign/DeclareDesign@rewrite", build_vignettes = TRUE)
```

`randomizr` from CRAN is enough. `vignette("getting-started")` is the short introduction; this vignette is the long one.

## What does not change

Here is a standard two-arm design, written identically in 1.x and 2.0.

```{r}
library(DeclareDesign)
library(fabricatr)
library(randomizr)
library(estimatr)

design <-
  declare_model(N = 600, U = rnorm(N), potential_outcomes(Y ~ 0.3 * Z + U)) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, .method = lm_robust, inquiry = "ATE", label = "ols")

diagnose_design(design, sims = 500)
```

Everything about that block carries over: the `+` operator, expressions that see the columns declared earlier in the same step, `N` available inside the data mask, `potential_outcomes()` and `reveal_outcomes()`, formula-first estimators with `.method =` and `.summary =`, `term =`, and the `inquiry` argument that links an estimator to its estimand. Writing the potential outcomes out by hand (`Y_Z_0 = U, Y_Z_1 = U + 0.3`) and revealing them with an arithmetic expression works just as well. `draw_data()`, `draw_estimands()`, `draw_estimates()`, `run_design()`, `simulate_design()`, `diagnose_design()`, `diagnose_designs()`, `reshape_diagnosis()`, `tidy()`, `get_diagnosands()`, `get_simulations()`, `declare_test()`, `declare_step()`, `declare_diagnosands()`, `set_diagnosands()` and `label_estimator()` all keep their names and what they return, and the book's designs are the evidence.

The book's way of redesigning is the 2.0 way. A value named above the design is a parameter of it:

```{r}
tau <- 0.3
design <-
  declare_model(N = 600, U = rnorm(N), Y_Z_0 = U, Y_Z_1 = U + tau) + ...

designs <- redesign(design, tau = c(0.1, 0.3, 0.5))
diagnose_design(designs, sims = 500)
```

`tau` is a name the model step reads. `redesign()` finds it, rebinds it in a copy of the environment the expression was written in, and the parameter travels forward: it appears as a column in the simulations table and as a grouping variable in the diagnosands table, so custom diagnosands can refer to it directly. A designer function's argument is the same thing, so `redesign(designer(), tau = c(0.1, 0.3))` sweeps a designer, and `expand_design()` is deprecated in its favour.

Printing a design shows the calls that declared it, as in 1.x, then the names `redesign()` can reach, with the value each currently holds, its kind, and the steps that would respond to a change. Asking for a change to a name that is not one of them warns instead of silently returning the design you started with:

```{r}
print(design)
#> Research design with 5 steps
#>
#> Step 1 (model): declare_model(N = 600, U = rnorm(N), Y_Z_0 = U, Y_Z_1 = U + tau)
#> Step 2 (inquiry): declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))
#> Step 3 (assignment): declare_assignment(Z = complete_ra(N))
#> Step 4 (measurement): declare_measurement(Y = reveal_outcomes(Y ~ Z))
#> Step 5 (estimator): declare_estimator(Y ~ Z, .method = lm_robust, inquiry = "ATE", label = "ols")
#>
#> Parameters and objects the design refers to:
#>  name value   kind declared steps
#>   tau   0.3 scalar    FALSE     1

redesign(design, taw = 0.5)
#> Warning: You requested a change to taw but taw is not found in the design.
```

## What changes, and why

### In DeclareDesign 2.0

#### `redesign()` reaches a name the design reads, not a number it writes down

`declare_model(N = 500)` puts 500 in the design. Nothing outside the design holds that number and nothing names it, so `redesign(design, N = 1000)` has nothing to change. In 1.x that call warned that `N` was not found and returned the design unchanged, at N = 500, which is the quiet kind of failure: three "designs" that are one design. In 2.0 it errors and says what to do:

```{r}
design <- declare_model(N = 500, U = rnorm(N), Y = U) + declare_inquiry(m = mean(Y))
redesign(design, N = 1000)
#> Error:
#> ! N is an argument this design writes down, not a parameter.
#> ℹ To redesign over values of `N`, give it a name outside the design:
#>   `N <- 500`, then `declare_model(N = N, ...)`. A designer function's
#>   argument does the same thing.
#> ℹ `declare_parameters(N = 500)` names it inside the design instead, which is
#>   worth it when several steps read it or when a column shares its name.
```

The ordinary way out is the book's: a name outside the design, or a designer function's argument. `declare_parameters(N = 500) + declare_model(N = N, ...)` does the same from inside the design, and earns its place when several steps read the value or when a column shares its name. A value written inline inside a call, as in `complete_ra(N, prob_each = c(0.2, 0.8))`, cannot be redesigned either: there is no name to rebind, so name it first. Functions count as names, so which estimator or which summary function a design uses can itself be a parameter: `redesign(design, g = median)` swaps `g` everywhere it appears.

A bare atomic vector always supplies one value per design, which is what makes `tau = c(0.1, 0.3)` two designs. A parameter that is itself a vector therefore has to be wrapped, in 2.0 as in 1.x, and 2.0 says so where 1.x handed back the broken designs without comment:

```{r}
redesign(design, prob_each = c(0, 0.5, 0.5))
#> Warning: `prob_each` currently holds 3 values, so c(0, 0.5, 0.5) is being
#> read as 3 designs, one value each.
#> ℹ Wrap it in `list()` to use it as a single replacement.

redesign(design, prob_each = list(c(0, 0.5, 0.5)))
```

The rule applies to bare vectors and bare lists, and to nothing else. A data frame, a matrix, a function and anything else carrying a class is one value, so a design built on a supplied data set is re-run on another one by naming it: `declare_model(data = pilot, ...)` is redesigned with `redesign(design, pilot = full_sample)`, reaching the name the design reads the data under rather than `data`, which names fabricate's argument. 1.x read the data frame as a list of columns and asked for one design per column.

#### `inquiry =` is the quoted label of the inquiry

```{r}
# old, sometimes seen
my_inquiry <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))
declare_estimator(Y ~ Z, inquiry = my_inquiry)
#> Error: Pass the inquiry label as a string, not a step object.

# new
declare_estimator(Y ~ Z, inquiry = "ATE")
```

In 1.x the step object carried a `label` attribute whose value was the step type, so passing it linked the estimator to an inquiry called `"inquiry"`, which does not exist, and the estimand column came back all `NA` with no complaint. Requiring the quoted label puts the link where a reader can check it. A label that matches no inquiry step now warns, once, naming the estimator, the label it asked for and the labels that exist; in 1.x it was silent.

#### `make_groups = vars(...)` is gone; group with `group_by()`

```{r}
# old
diagnose_design(design, make_groups = vars(N))
#> Error: `make_groups` is not a design.
#> ℹ `make_groups` is gone. Group the simulations instead:
#>   `simulate_design(design) |> group_by(N) |> diagnose_design()`.

# new
simulate_design(design, sims = 500) |>
  group_by(N) |>
  diagnose_design()
```

`diagnose_design()` accepts a simulations data frame as its first argument and honours any grouping already on it. Anything you can compute with `mutate()` becomes a grouping variable, so subgroup diagnosis no longer needs a bespoke argument:

```{r}
simulate_design(design, sims = 500) |>
  mutate(significant = p.value < 0.05) |>
  group_by(significant) |>
  diagnose_design()
```

You can also filter simulations before diagnosing, bind simulations from separate runs, or diagnose a table you saved to disk last week. `add_grouping_variables` and `future.seed` are gone the same way, and each errors with its replacement: grouping is `group_by()`, and seeding is `set.seed()` before the call under any plan (see [Parallel execution]).

#### Argument names that moved out of the way of your parameters

`redesign()` takes `.design`, with `.expand` after the dots. Everything else in those calls is a parameter of *your* design, and a plain `design` partially matched a parameter named `d`, `de` or `des` and swallowed it; `.method` and `.summary` in `declare_estimator()` follow the same convention. `redesign(design = d, N = 10)` therefore errors; write `redesign(d, N = 10)`. The designs a sweep returns are named `design_1`, `design_2` and so on.

`label_estimator(fn = )` and `label_test(fn = )` are `label_estimator(.method = )`; `set_diagnosands(x = )` is `set_diagnosands(design = )`. Each is an error if written the old way.

#### `draw_data()` takes the design and nothing else

`draw_data(design, data = , start = , end = )` ran a slice of the design against supplied data. That job belongs to `get_estimates(design, data)` for the estimators, and to `design[start:end]` for a slice, since a design is a list of steps and `+` reassembles one. `draw_data()` now runs the data-generating steps only, so a design whose estimator fails still draws its data, and drawing data no longer fits models it throws away.

#### `reshape_diagnosis()` reshapes; select columns afterwards

`reshape_diagnosis(select = , exclude = )` are gone. The result is a tibble; `select()` it.

#### `expand_design()` is deprecated

A designer's argument is a name the design reads, so `redesign(designer(), N = c(100, 200))` does what `expand_design(designer, N = c(100, 200))` did, with one verb instead of two. `expand_design()` still works and warns once per session. The one thing it did that `redesign()` cannot is rebuild a design whose shape depends on an argument (a designer that adds a step when `blocked = TRUE`); call the designer with each value for that.

#### Step-modification functions are deprecated

`insert_step()`, `delete_step()` and `replace_step()` still work, with their 1.x argument order, and warn once per session. The replacement is to rebuild the design, which is available because a design is an ordinary list of steps:

```{r}
design_v2 <- design[1:2] + new_step + design[4:5]
```

#### The code-comparison functions are removed; `compare_diagnoses()` is not

`compare_designs()`, `compare_design_code()`, `compare_design_summaries()` and `print_code()` error with a message asking for the R script instead. Reconstructing a legible declaration from a design object requires carrying the original source text through every transformation the design undergoes, and the reconstruction was never faithful enough to trust. Sharing the script that made the design is simpler.

Comparing what two designs *achieve* is a different question, and `compare_diagnoses()` answers it:

```{r}
compare_diagnoses(design, redesign(design, tau = 0.5), sims = 500)
```

Every diagnosand the two designs share is reported under each design and as a difference, with a bootstrap standard error and interval on that difference. Either argument can be a diagnosis you already ran.

#### `declare_potential_outcomes()`, `declare_reveal()` and `declare_estimators()` are gone

The first two were already deprecated in 1.x and error with the replacement: `potential_outcomes()` inside `declare_model()` and `reveal_outcomes()` inside `declare_measurement()`. `declare_estimators()` was an alias of `declare_estimator()` and is simply absent. `declare_population()` still works and warns once.

#### What comes back has one more column, in a different order

Every table of estimates (`draw_estimates()`, `run_design()`, `get_estimates()`, the simulations) carries an `estimand` column, joined from the inquiry, and puts `estimator`, `inquiry` and `estimand` after the tidied statistics rather than before. Code that indexes columns by position, or that snapshots a table, will see the difference; code that uses names will not. Two estimators with the same label are renamed apart with a message where 1.x errored.

#### An estimator that fails on one draw is recorded, not fatal

Under `simulate_design()` and `diagnose_design()` a draw whose estimator errors is kept as a row with `error = TRUE` and the message, one warning per run gives the count per estimator and the first message, and diagnosands are computed on the draws that succeeded, with `n_sims` reporting how many that was. In 1.x one non-convergence in five hundred lost the run. `run_design()` and `draw_estimates()` still stop, because a single run is somebody debugging.

### In fabricatr 2.0

fabricatr 2.0 has its own vignette, `vignette("fabricatr2.0", package = "fabricatr")`, and its changes are confined to the spelling of multilevel and cross-classified declarations. A flat `declare_model(N = 500, ...)` is unaffected, and so is the output of every declaration that runs: the same columns, in the same order, with the same types, `ID` still zero-padded character, `N` still available in the data mask and still absent from the output.

The spellings that change, each accepted for now with a once-per-session warning that prints the call to write instead: `add_level(nest = FALSE)` becomes `declare_level()`, `cross_levels(by = join_using(A, B))` and `link_levels(by = join_using(A, B, rho = ))` become `.by = c("A", "B")` with `rho` as its own argument, and `modify_level(by = )` becomes `modify_level(.by = )`. `fabricate(100, ...)` with a positional `N` is an error; write `N = 100`.

## Porting an existing script

Seven greps cover every breaking change in both packages. Run them over a script and whatever they do not find needs no attention.

| Grep for | Replace with | Where it bites |
|---|---|---|
| `redesign(` over a number written inside a step | name the value above the design, or `declare_parameters()` | every redesign of a literal |
| `inquiry = ` followed by an object | `inquiry = "ATE"` | any estimator linked to an estimand |
| `make_groups`, `add_grouping_variables` | `simulate_design() \|> group_by() \|> diagnose_design()` | subgroup diagnosis |
| `redesign(design = `, `expand_design(` | `redesign(d, ...)`, `redesign(designer(), ...)` | named first argument; deprecated verb |
| `draw_data(` with `data =`, `start =` or `end =` | `get_estimates(design, data)`, `design[i:j]` | partial runs |
| `nest = FALSE`, `join_using`, `modify_level(by =` | `declare_level()`, `.by = c("A", "B")`, `.by =` | multilevel fabrication |
| `fabricate(<number>,` | `fabricate(N = <number>,` | positional `N` |

Every row fails loudly. The DeclareDesign rows stop with an error naming the replacement, and the fabricatr rows warn once and run, which is the property that matters when you are porting someone else's code and cannot check every number by hand.

Two things do not fail at all, and are the ones to check by eye: the `estimand` column and the column order of every estimates table, and the deprecated verbs (`expand_design()`, `insert_step()`, `delete_step()`, `replace_step()`), which run with a warning. `compare_designs()` and `print_code()` stop outright.

## What is new

### `declare_parameters()` and `declare_notes()`

A parameter is a value a caller may set; a note is a quantity the design works out. `declare_parameters()` names the first, `declare_notes()` the second, and `redesign()` reaches only the first.

```{r}
design <-
  declare_parameters(m_arms = 3, n_units = 300) +
  declare_notes(conditions = seq_len(m_arms)) +
  declare_model(N = n_units, U = rnorm(N)) +
  declare_assignment(Z = complete_ra(N, conditions = conditions)) +
  ...

design_parameters(design)
redesign(design, m_arms = c(2, 3, 4))
```

A declared parameter is bound for every step after it and takes precedence over a workspace binding of the same name, including one read through a helper function. It is evaluated once, when the design is built, so `declare_parameters(u = rnorm(1))` fixes one number for the life of the design; put `set.seed()` above the declaration, not above `draw_data()`, if that number is to be reproducible. A note is evaluated at its own position on every run, so it can read the data (`declare_notes(tallest = max(Y))` after a model step) and it survives a sampling step that destroys what it was computed from.

### A design is a value

A design carries the objects its expressions read: helper functions, data frames, the `tau` above. It therefore survives `saveRDS()` and a fresh session, reaches a parallel worker, and does not change meaning when an unrelated console variable is edited. 1.x designs were live views onto the environment they were written in, which is what made issue #293 possible.

### Step-level `draws`

Every step takes a `draws` argument. When a step declares `draws = k`, it fans out `k` independent times, and every step downstream inherits and multiplies those paths.

```{r}
design <-
  declare_model(N = 500, potential_outcomes(Y ~ tau * Z + rnorm(N)), draws = 100) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N), draws = 50) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE")

# 100 populations x 50 assignments within each = 5,000 simulation paths
diagnosis <- diagnose_design(design)
```

The declared draws define the simulation plan, so no `sims` argument is needed. The simulations table gains one column per fan-out step: `model_draw` runs 1 to 100 and `assignment_draw` runs 1 to 50 within each. Each population carries its own realized estimand, and the estimates drawn within it are matched to that estimand rather than to a pooled average. Steps upstream of the first fan-out run once and stay fixed, which is what "hold the population fixed and redraw the assignment" means.

### Variance decomposition

When a design has declared draws, the diagnosis carries an extra slot:

```{r}
diagnosis$variance_decomposition
```

Each row is one per-simulation quantity (`estimate`, `p.value`, `std.error`, `statistic` by default) and each column is one draw level, named after the step that produced it. `var_model` is the between-population variance, `var_assignment` is the variance across randomizations within a fixed population, and `var_residual` is what remains inside a single (population, assignment) cell. The `frac_` columns normalize the components so they sum to 1: `frac_model` of 0.79 says that 79 percent of the variability in the estimate traces to which population was drawn rather than to which randomization was realized.

### Parallel execution

```{r}
future::plan(future::multisession, workers = 4)
simulate_design(design, sims = 2000)
future::plan(future::sequential)
```

No other change is needed. Every draw runs on its own L'Ecuyer-CMRG random-number stream whichever plan is active, so `set.seed(1)` before the call gives the same table under `plan(sequential)` and `plan(multisession)`, nested draws included. The packages attached in your session are attached on each worker. The seeds are not 1.x's, so a seed carried over from a 1.x script reproduces a 1.x table only under 1.x. The bootstrap inside `diagnose_design()` runs in the main process; it is one grouped `summarize()` over the stacked replicates and is not where the time goes.

### `summary()` says what each step did

`summary(design)` runs the design once and reports, under each step's call, what that step did to the data: the rows a data step produced or kept and the columns it added, dropped or changed, the value each inquiry took, and the estimate each estimator returned, followed by the run itself and the parameters. It replaces 1.x's variable-by-variable tables with one line per step, and the pieces are on the returned object (`summary(design)$steps`, `$estimates`, `$data`) rather than only on the console. `summary(design, run = FALSE)` skips the run for a design whose estimator is slow.

```{r}
summary(design)
#> Step 3 (sampling): declare_sampling(S = complete_rs(N, n = 50))
#>   keeps 50 of 100 rows; adds S
#> Step 4 (assignment): declare_assignment(Z = complete_ra(N))
#>   adds Z
```

### Progress

`simulate_design(design, progress = TRUE)` shows a bar for that call. The better habit is `progressr::handlers(global = TRUE)` once per session, after which every call reports, in the console, in a Shiny app or from inside parallel workers, through whatever handler you chose. The package signals progress and never renders it, so with no handler it prints nothing.

### Estimator arguments pass as written

`declare_estimator()` and `declare_step()` hand their arguments to `.method` or the handler as written rather than pre-evaluated against the data, so a method that does its own non-standard evaluation gets what it expects: `metafor::rma.uni` (issue #463), `tidyr::pivot_wider` as a `declare_step()` handler with `id_cols = pair`, and a helper that reads a bare column name. A plain function taking values works as before, because `.method` is called with the data and the expressions are evaluated where the method evaluates them.

## Speed

All timings are medians of five runs on an Apple M4 Pro under R 4.6.0, comparing DeclareDesign 1.1.1 with fabricatr 1.0.2 against DeclareDesign 2.0.0 with fabricatr 2.0.0, on the two-arm design declared above (N = 600, `lm_robust`), measured on the 2026-08-24 build.

| Operation | 1.1.1 | 2.0.0 | Ratio |
|---|---:|---:|---:|
| `simulate_design(sims = 500)` | 0.97 s | 0.84 s | 1.15x |
| `diagnose_design(sims = 500, bootstrap_sims = 200)` | 1.29 s | 0.96 s | 1.34x |
| `simulate_design(sims = 2000)`, four workers | | 1.07 s against 3.37 s sequential | 3.2x |

Parallelism is the larger lever. The sequential gain is concentrated in the plumbing, because model fitting is the same `lm_robust()` call in both packages and takes more than half of each simulation; on this design the estimator step is about 1 ms of a 1.7 ms draw. The bootstrap stacks all `B` resampled tables with a `.boot_id` column and runs one grouped `summarize()` instead of `B` separate calls, which is where the diagnosis row's extra gain comes from. Speeding up the estimator itself is what estimatr 2.0 is for.

On a clustered design (30 villages of 20 citizens, ICC 0.2, cluster-robust standard errors) the two packages agree on what they report, which is the check worth running whenever a rewrite claims to be a replacement. Across 500 simulations: bias -0.010 against 0.015, SD of the estimate 0.197 against 0.194, power 0.294 against 0.304, coverage 0.968 against 0.954.

## Why rewrite at all

The original packages carry machinery solving problems the tidyverse already solves.

fabricatr's workspace system exists to answer one question: how does a column defined early in a `fabricate()` call become visible to a column defined later? Its answer is a UUID-keyed environment of data frames, an "active df" pointer, `rlang::call_modify` injection, and `dynGet("N")` walking the dynamic scope to find the row count. The tidyverse answer is a named list used as a data mask and a loop over `rlang::eval_tidy`. fabricatr 2.0 is about 2,100 lines of R against 3,193.

DeclareDesign's complexity concentrates in three places, and each has a tidyverse replacement. The simulation loop is `purrr::map2`, or `furrr::future_map2` when a parallel plan is set, over one RNG stream per draw. Diagnosis is one `group_by() |> summarize()`. Parameter substitution in `redesign()` is environment cloning: a quosure already carries the environment it was captured in, so changing a parameter is cloning that environment and rebinding one name, and the same operation is what `declare_parameters()` does at construction and `declare_notes()` does at run time. A `design_step` is a `function(data)` with attributes, and a `design` is a named list of them.

DeclareDesign 2.0 is 6,153 lines of R against 1.1.1's 7,036, so the line count is not the argument. The argument is what the lines are spent on. In 1.x the count goes to the custom evaluation machinery; in 2.0 it goes to capturing what a design reads so that it is a value, to the checks that turn silent failures into messages, and to the features under [What is new]. The remaining code is written in idioms a reader already knows from dplyr, purrr and rlang, which is the difference between a package two people can maintain and a package two people can hand to someone else.

## Current status

The rewrite is pre-release. Nothing goes to CRAN until fabricatr 2.0 and estimatr 2.0 are accepted, since DeclareDesign depends on both.

**Coverage the claims rest on.** 875 assertions in the package's own suite, 0 failures. The 81 book declarations are `tests/testthat/test-book-designs.R`, each asserting that the declaration builds, runs, and returns estimates that are not all `NA`; the file skips on CRAN because it needs a dozen modelling packages. All 191 code chunks of the book's fifteen code-bearing chapters have been run in order under 2.0: 173 ran without error, two are `install.packages()` calls, one chapter halts where it attaches the CRAN DesignLibrary over 2.0, and the remaining errors are names the book defines outside any displayed chunk. The designs of `macartan/ResearchDesigns` run under 2.0, 494 assertions with 0 failures, and a no-op redesign of each of their 194 parameters changes nothing. `R CMD check --as-cran` is clean apart from one NOTE for the `Remotes:` field, which stays until the dependencies ship.

**Of the nine book declarations not checked in**, four use a value the book defines outside any code chunk, one needs the `interference` package, which is not on CRAN, one fails identically under 1.1.1, and three passed tidyselect arguments through `declare_step()`, which now passes arguments as written; they have not been re-added to the suite.

**Fixed in the 2026-08-24 review**, each with a regression test: a declared parameter read through a helper function was ignored; `label_estimator()` of a workspace function failed on every parallel draw; sequential and parallel runs gave different numbers for the same seed; `make_groups` and `future.seed` were accepted and silently dropped; a misspelled `inquiry` label was silent; a sampling step with no `S` column kept every row without a word; an error inside a step did not say which step; `insert_step()` had its `before` and `after` arguments in the wrong order.

**Open.** `library(DesignLibrary)` attaches CRAN DeclareDesign 1.1.1 over 2.0 until DesignLibrary is released against 2.0, and the result is a hybrid that fails on the next declaration.

### Open issues in DeclareDesign

Every open issue on the DeclareDesign tracker was run against both packages. The script is in the repository, so this table can be regenerated rather than trusted.

**Seven close.** Each of these errors or misbehaves in DeclareDesign 1.1.1 and works here.

| Issue | What breaks in DeclareDesign 1.1.1 |
|---|---|
| [#463](https://github.com/DeclareDesign/DeclareDesign/issues/463) | `rma.uni` as the estimator's method: quosures reach a function that uses base non-standard evaluation |
| [#456](https://github.com/DeclareDesign/DeclareDesign/issues/456) | `declare_estimator(x ~ 1, model = lm, subset = x > 1)` errors, though it works with `lm_robust` |
| [#456](https://github.com/DeclareDesign/DeclareDesign/issues/456) | the same with `weights = 1/x^2` |
| [#457](https://github.com/DeclareDesign/DeclareDesign/issues/457) | a handler's own default arguments are not picked up |
| [#479](https://github.com/DeclareDesign/DeclareDesign/issues/479) | a spurious many-to-many merge warning when two estimators name the same inquiry |
| [#482](https://github.com/DeclareDesign/DeclareDesign/issues/482) | a design with a test but no estimator cannot be diagnosed after `redesign()` |
| [#509](https://github.com/DeclareDesign/DeclareDesign/issues/509) | a recursive helper function in the calling environment overflows the node stack |

Six of the seven are the same cause seen from different angles. 1.x splices an estimator's arguments into the method call as quosures, so every method not written for rlang receives an object it cannot use.

**Two feature requests are now implemented.** [#385](https://github.com/DeclareDesign/DeclareDesign/issues/385) asked for an estimator that errors to produce a row rather than abort the simulation; see [An estimator that fails on one draw is recorded, not fatal]. [#293](https://github.com/DeclareDesign/DeclareDesign/issues/293), a helper function whose environment holds a parameter surviving deletion of that parameter, follows from a design being a value:

```{r}
m <- 2
f <- function(x) m * x
design <- declare_model(N = 2, X1 = f(1)) + NULL
rm(m)
draw_data(design)
```

**One is not.** [#497](https://github.com/DeclareDesign/DeclareDesign/issues/497) asks for `draw_data(design, theta = 0.5)`; both packages go through `redesign()`.
