---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(343)
```

```{r, eval = FALSE}
library(DeclareDesign)
library(fabricatr)
library(randomizr)
library(estimatr)
```

```{r, include = FALSE}
# TEMPORARY, delete at release. The 2.0 packages are not yet on CRAN under
# their own names, so the chunks below run against the branch packages while
# the chunk above shows what a reader will type.
library(DeclareDesign)
library(fabricatr)
library(randomizr)
library(estimatr)
```

`DeclareDesign` declares and diagnoses. The other three supply the pieces a
declaration is written out of: `fabricatr` makes data, `randomizr` randomizes,
and `estimatr` estimates.

## Declaration

Every design element is made with a function that starts with the word `declare`. Here is one, an assignment that gives each unit a 60 percent chance of treatment:

```{r}
simple_random_assignment <- declare_assignment(Z = simple_ra(N = N, prob = 0.6))
```

Design elements are themselves functions, which is why the same declaration produces a different assignment each time you use it rather than one fixed vector.

A two-arm randomized experiment with 100 units puts four elements together with `+`. `declare_parameters()` names the numbers we intend to vary later; everything else is written down as part of the design:

```{r}
declaration <-
  declare_parameters(n_units = 100) +
  declare_model(
    N = n_units,
    potential_outcomes(Y ~ 0.2 * Z + rnorm(N))
  ) +
  declare_inquiry(ATE = mean(Y_Z_1) - mean(Y_Z_0)) +
  declare_assignment(Z = complete_ra(N)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, .method = difference_in_means, inquiry = "ATE")
```

Run the design to see what one study would look like:

```{r}
head(draw_data(declaration))
```

Both potential outcomes are in the data because you are simulating the world. A real study observes `Y` alone.

```{r}
draw_estimands(declaration)

draw_estimates(declaration)
```

## Diagnosis

Diagnosis is the process of simulating the design many times and calculating summary statistics about the design that describe its properties, which we call diagnosands.

```{r}
diagnosis <- diagnose_design(declaration, sims = 100)

diagnosis
```

Bias sits near zero, so the estimator recovers the estimand on average. Power is the number to look at here: with 100 units and an effect of 0.2, most runs of this study end without a significant result.

Each diagnosand comes with a bootstrapped standard error, so you can tell a real difference from simulation noise. `?default_diagnosands` says what each one means.

## Redesign

We redesign to learn how the diagnosands change as design features change. `redesign()` over a range of sample sizes produces a list of designs:

```{r}
designs <- redesign(declaration, n_units = c(100, 200, 300, 400, 500))

diagnose_design(designs, sims = 100)
```

A design can be redesigned over the objects its expressions read from outside it and over the parameters it declares. A number written into the design, as `0.2` is above, is part of the design: to redesign over values of it, give it a name first, either outside the design or in `declare_parameters()`.

Power rises with the sample size while bias stays where it was. Choosing a sample size is now an argument about a table rather than a hunch, and none of it required collecting data.

## A design is code

The declaration above is the whole design: the model, the inquiry, the data strategy and the answer strategy, in one object you can run, diagnose, change, and send to a collaborator. `cite_design()` gives the citation, `print_code()` gives the code back, and `?DeclareDesign` lists the verbs.

The book at <https://book.declaredesign.org> works through the same steps at length, and its library of designs covers the common ones.
