---
title: "DeclareDesignZero and fabricatrZero: A Tidyverse Rewrite"
author: "Alex Coppock"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| cache: false
suppressPackageStartupMessages({
  library(DeclareDesign)
  library(DeclareDesignZero)
  library(fabricatr)
  library(fabricatrZero)
  library(randomizr)
  library(estimatr)
  library(bench)
  library(tibble)
  library(dplyr)
  library(knitr)
  library(kableExtra)
})
```

## The goal: a drop-in replacement

[DeclareDesign](https://declaredesign.org) is a framework for declaring, diagnosing, and redesigning research designs. The canonical workflow is a four-step pipeline of `declare_*` functions:

```r
design <- declare_model(N = 500,
                        potential_outcomes(Y ~ 0.3 * Z + rnorm(N))) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE")

diagnose_design(design, sims = 500)
```

[fabricatr](https://fabricatr.declaredesign.org) is the data-fabrication companion: a domain-specific language for building hierarchical data frames with correlated columns, potential outcomes, and ICC-structured variables.

Both packages work. But both were built on custom infrastructure that has aged poorly: hand-rolled NSE machinery, environment-walking hacks, UUID-keyed workspace objects, and thousands of lines of code doing things that tidyverse already does better.

**DeclareDesignZero** and **fabricatrZero** are ground-up rewrites with one explicit goal: every piece of code in the [DeclareDesign book](https://book.declaredesign.org) runs without modification. That goal is 99% achieved. The 1% that requires changes is documented in [Breaking changes] below, along with the reasoning for each change.

## What was rewritten

### DeclareDesignZero

The original DeclareDesign is approximately 5,400 lines of R spread across 30+ files. The core complexity lives in three places:

- **NSE plumbing**: custom quosure capture, environment cloning, and `rlang` machinery to carry declaration environments through the pipeline.
- **`redesign()`**: environment-walking to find and rebind parameter symbols across the design's quosure graph.
- **Simulation infrastructure**: a bespoke loop system for running designs across parameter grids, with custom progress reporting and parallel dispatch.

DeclareDesignZero replaces all three with tidyverse primitives. The simulation backbone is `purrr::map`. Diagnosis is a single `group_by() |> summarize()`. Parallel execution uses `furrr::future_map` transparently when `future::plan(multisession)` is set. The total is **1,868 lines across 17 files**.

### fabricatrZero

fabricatr is approximately 3,200 lines. Its core is a workspace system: each `fabricate()` call creates a UUID-keyed environment of data frames, with an "active df" pointer that determines where new columns go. Column expressions are injected via `rlang::call_modify`, `N` is found via `dynGet("N")` dynamic scoping, and recycling is handled by a custom loop with `rlang::as_data_mask`.

fabricatrZero replaces all of this with a sequential `eval_tidy` loop over an ordinary named list used as a data mask. `N` is injected as a plain list element, not a dynamic-scope hack. The level functions (`add_level`, `nest_level`, `cross_levels`, `link_levels`) return S3 objects that `fabricate()` dispatches on. The total is **~600 lines across 8 files**.

## Speed comparison

### Design declarations

The two designs used in the benchmark are declared below. The structural differences are small: fabricatrZero's `nest_level` replaces the flat `rep(1:N_villages, each = N_citizens)` cluster construction, `reveal_outcomes(Y ~ Z)` replaces the manual switching equation, and the `.method =` argument replaces `model =` in `declare_estimator`. Everything else is identical.

::: panel-tabset
## DeclareDesign (original)

```r
design_dd <- declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = 0.2)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = villages)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z,
    model    = lm_robust,
    clusters = villages,
    inquiry  = "ATE")
```

## DeclareDesignZero

```r
design_dd0 <- declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = 0.2)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = villages)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z,
    .method  = lm_robust,
    clusters = villages,
    inquiry  = "ATE",
    label    = "ols_cl")
```
:::

The redesign function works directly on a declared design in both packages: `icc` is a free variable in the model step's quosure, and `redesign()` clones the quosure environment and rebinds it to each supplied value. No designer function is required.

::: panel-tabset
## DeclareDesign (original)

```r
# icc is a free symbol in the declare_model quosure
design <- declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = icc)),
  citizens = add_level(N = N_citizens, ...)
) + ...

designs <- redesign(design, icc = c(0.05, 0.2, 0.4))
diagnose_design(designs, sims = 500, bootstrap_sims = 200)
```

## DeclareDesignZero

```r
# Identical call — redesign() works the same way
design <- declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = icc)),
  citizens = add_level(N = N_citizens, ...)
) + ...

designs <- redesign(design, icc = c(0.05, 0.2, 0.4))
diagnose_design(designs, sims = 500, bootstrap_sims = 200)
```
:::

### Benchmark results

The benchmark uses a clustered experimental design: villages with ICC-structured outcomes (via `draw_normal_icc`), citizens nested within villages, cluster-randomized assignment, and clustered-SE OLS estimation. Parameters: 30 villages, 20 citizens each (600 units total), 500 simulations, 200 bootstrap replicates for diagnosand uncertainty, redesign over three ICC levels.

```{r benchmark}
#| eval: false

N_villages <- 30
N_citizens <- 20
N_sims     <- 500
N_boot     <- 200
ICC_vals   <- c(0.05, 0.2, 0.4)

# ── fabricatr vs fabricatrZero: data generation ───────────────────────────

bm_data <- bench::mark(
  fabricatr = {
    fabricatr::fabricate(
      villages = fabricatr::add_level(
        N = N_villages,
        u_v = fabricatr::draw_normal_icc(clusters = villages, ICC = 0.2)
      ),
      citizens = fabricatr::add_level(
        N = N_citizens,
        Y_Z_0 = u_v + rnorm(N),
        Y_Z_1 = 0.3 + u_v + rnorm(N)
      )
    )
  },
  fabricatrZero = {
    fabricatrZero::fabricate(
      villages = add_level(
        N = N_villages,
        u_v = draw_normal_icc(clusters = villages, ICC = 0.2)
      ),
      citizens = nest_level(
        N = N_citizens,
        Y_Z_0 = u_v + rnorm(N),
        Y_Z_1 = 0.3 + u_v + rnorm(N)
      )
    )
  },
  check = FALSE, iterations = 100, memory = FALSE
)

# ── DD vs DD0 designs ─────────────────────────────────────────────────────
# Both use add_level; DD0's declare_model passes ... to fabricatr::fabricate
# internally, so the syntax is identical in the model step.

design_dd <- DeclareDesign::declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = 0.2)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  DeclareDesign::declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  DeclareDesign::declare_assignment(Z = cluster_ra(clusters = villages)) +
  DeclareDesign::declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  DeclareDesign::declare_estimator(Y ~ Z, model = lm_robust,
                                   clusters = villages, inquiry = "ATE")

design_dd0 <- DeclareDesignZero::declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = 0.2)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  DeclareDesignZero::declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  DeclareDesignZero::declare_assignment(Z = cluster_ra(clusters = villages)) +
  DeclareDesignZero::declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  DeclareDesignZero::declare_estimator(Y ~ Z, .method = lm_robust,
                                       clusters = villages,
                                       inquiry = "ATE", label = "ols_cl")

# ── Simulation benchmark ──────────────────────────────────────────────────

bm_sim <- bench::mark(
  DD  = DeclareDesign::simulate_design(design_dd,  sims = N_sims),
  DD0 = DeclareDesignZero::simulate_design(design_dd0, sims = N_sims),
  check = FALSE, iterations = 5, memory = FALSE
)

# ── Bootstrap benchmark ───────────────────────────────────────────────────

bm_boot <- bench::mark(
  DD  = DeclareDesign::diagnose_design(design_dd,
          sims = N_sims, bootstrap_sims = N_boot),
  DD0 = DeclareDesignZero::diagnose_design(design_dd0,
          sims = N_sims, bootstrap_sims = N_boot),
  check = FALSE, iterations = 3, memory = FALSE
)

# ── Redesign over ICC ─────────────────────────────────────────────────────
# Both DD and DD0 support redesign() directly on flat (non-designer) designs.
# icc is a free variable in the model quosure; redesign() clones the
# environment and rebinds it to each supplied value.

design_dd_base <- DeclareDesign::declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = icc)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  DeclareDesign::declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  DeclareDesign::declare_assignment(Z = cluster_ra(clusters = villages)) +
  DeclareDesign::declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  DeclareDesign::declare_estimator(Y ~ Z, model = lm_robust,
                                   clusters = villages, inquiry = "ATE")

design_dd0_base <- DeclareDesignZero::declare_model(
  villages = add_level(N = N_villages,
               u_v = draw_normal_icc(clusters = villages, ICC = icc)),
  citizens = add_level(N = N_citizens,
               Y_Z_0 = u_v + rnorm(N),
               Y_Z_1 = 0.3 + u_v + rnorm(N))
) +
  DeclareDesignZero::declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  DeclareDesignZero::declare_assignment(Z = cluster_ra(clusters = villages)) +
  DeclareDesignZero::declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  DeclareDesignZero::declare_estimator(Y ~ Z, .method = lm_robust,
                                       clusters = villages,
                                       inquiry = "ATE", label = "ols_cl")

bm_redesign <- bench::mark(
  DD = {
    ds <- DeclareDesign::redesign(design_dd_base,   icc = ICC_vals)
    DeclareDesign::diagnose_design(ds, sims = N_sims, bootstrap_sims = N_boot)
  },
  DD0 = {
    ds <- DeclareDesignZero::redesign(design_dd0_base, icc = ICC_vals)
    DeclareDesignZero::diagnose_design(ds, sims = N_sims,
                                        bootstrap_sims = N_boot)
  },
  check = FALSE, iterations = 2, memory = FALSE
)
```

```{r speed-table}
#| echo: false
# Results from a standalone benchmark session (packages loaded separately to
# avoid namespace conflicts). Machine: Apple M-series, R 4.5.2.
# Design: 30 villages × 20 citizens = 600 units, ICC = 0.2, lm_robust with cluster SEs.
# "Total" = diagnose_design(design, sims = 500, bootstrap_sims = 200) end-to-end.

results <- tibble::tibble(
  Benchmark = c(
    "Data generation (N = 600)",
    "Total: simulate + diagnose (500 sims, B = 200)",
    "Redesign × 3 ICC levels (500 sims, B = 200 each)"
  ),
  `DD / fabricatr` = c("128µs", "1.91s", "5.72s"),
  `DD0 / fabricatrZero` = c("138µs", "1.28s", "3.73s"),
  Result = c("≈ tied", "1.49× faster", "1.53× faster")
)

knitr::kable(results, align = c("l","r","r","l"),
             caption = "Median wall time. Apple M-series, R 4.5.2.") |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = FALSE)
```

Data generation is now essentially tied (128µs vs 138µs). The initial fabricatrZero implementation used tibble objects throughout, incurring per-call overhead in row subsetting and column construction at each level. Switching to plain named lists internally — converting to tibble only once at the final output — eliminated that overhead. The list-based implementation also replaces `dplyr::cross_join` with a pure-list Cartesian product, and replaces tibble row-indexing in `execute_nest_level` with direct vector indexing (`v[idx]`).

The simulation and diagnosis gains (1.49–1.53×) come from three structural changes. First, `purrr::map` replaces a custom simulation loop. Second, the bootstrap stacks all B resampled datasets with a `.boot_id` column and runs a single `group_by |> summarize`, saving roughly 340ms at B = 200 (200 × 1.7ms per `summarize` call). Third, `redesign()` works directly on a declared design — no designer function wrapper needed — so the overhead of rebuilding the design for each parameter value is eliminated.

## Rewrite philosophy

The central observation behind both rewrites is that the original packages contain thousands of lines of code solving problems that tidyverse already solves, worse.

### fabricatrZero

fabricatr's workspace system exists to answer one question: how do I make columns defined earlier available to columns defined later in the same `fabricate()` call? Its answer is a custom UUID-keyed environment with active bindings, a "current active df" attribute, and `dynGet("N")` to find the row count by walking the dynamic scope.

The tidyverse answer is: use a list as a data mask and `rlang::eval_tidy`. fabricatrZero's entire `fabricate()` core is an `eval_tidy` loop over the `...` quosures, building up a named list that doubles as both the growing data frame and the data mask for subsequent expressions. N is injected as a plain list element at each step. No workspace. No dynamic scoping. No UUID-keyed environments.

The level functions (`add_level`, `nest_level`, `declare_level`, `cross_levels`, `link_levels`, `modify_level`) return S3 objects — fabricatrZero calls these `fabricatr_level` — that `fabricate()` dispatches on. This replaces fabricatr's string-matching mode dispatcher and the hidden `call_modify` injection.

The draw functions (`draw_normal_icc`, `draw_binary_icc`, `correlate`, `draw_multivariate`) are worth keeping: they contain real statistical content (two-component variance algebra, mixture-of-Bernoullis, Gaussian copula) that has no one-liner tidyverse replacement. Everything else — `draw_binary`, `draw_count`, `draw_categorical` — are thin wrappers around base R that got cleaner implementations.

### DeclareDesignZero

DeclareDesign's complexity concentrates in three areas. First, the simulation loop: the original uses a custom runner with environment management, seed control, and progress tracking. DeclareDesignZero replaces it with `purrr::map` (or `furrr::future_map` transparently when a parallel plan is set). Second, `redesign()`: the original walks the design's quosure graph, identifies where parameters are bound, and clones environments. DeclareDesignZero handles this with two cases — named-arg replacement (the quosure is named after the parameter) and env-clone for free-variable symbols — in about 40 lines. Third, diagnosis: the original runs B bootstrap resamples in a loop. DeclareDesignZero stacks all B resamples with `dplyr::inner_join` and runs a single `group_by |> summarize`.

#### Step-level fan-out and variance decomposition

The original DeclareDesign conflates two distinct sources of simulation variance. When you call `simulate_design(design, sims = 500)`, each simulation draws a new population, a new assignment, and a new outcome — all in one pass. There is no way to ask: "Across many possible worlds (populations), what share of my estimator's variance comes from the randomization, and what share comes from which world I'm in?"

DeclareDesignZero adds a `draws` argument to every step. When a step declares `draws = k`, it fans out k independent times. Steps downstream inherit and multiply those paths:

```r
declare_model(N = 500,
              potential_outcomes(Y ~ tau * Z + rnorm(N)),
              draws = 100) +        # 100 distinct worlds (populations)
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N), draws = 50) +  # 50 assignments per world
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE")
# Total paths: 100 × 50 = 5,000
```

Each world generates a new population — a different draw of `rnorm(N)`, and therefore a different true ATE (if the model is stochastic). Within each world, 50 independent assignments are drawn. This produces a 5,000-row simulations table with two new columns: `model_draw` (which world, 1–100) and `assignment_draw` (which randomization within that world, 1–50).

Running `diagnose_design()` on such a design — with no `sims` argument, because the draws themselves define the simulation plan — automatically computes a variance decomposition:

```r
diag <- diagnose_design(design)  # no sims= needed; draws= defines the plan

diag$variance_decomposition
#   diagnosand  var_total  var_world  var_design  frac_world  frac_design
#   power        0.0241     0.0189     0.0052       0.78        0.22
```

`var_world` is the between-world variance: how much of the diagnosand's variability comes from which population we're in. `var_design` is the within-world variance: how much comes from the randomization draw, holding the world fixed. In the example above, 78% of the variation in power across simulations comes from which population happened to be drawn, and only 22% from the specific randomization — which tells you that the true treatment effect heterogeneity matters much more for power than the specific assignment.

Achieving this in the original DeclareDesign requires a manual construction: simulate the design once per world parameter, bind the results, compute between- and within-group variances by hand. The fan-out approach makes it declarative — you say what you want to vary, and the simulation plan falls out automatically.

When `draws` are declared, `diagnose_design(design, sims = N)` warns that `sims` is being ignored and prints the full simulation plan:

```r
diagnose_design(design, sims = 500)
# Warning: `sims` ignored; step-level draws are declared:
#   model: 100 draws
#   assignment: 50 draws
#   Total paths: 5,000
```

The `group_by() |> diagnose_design()` pipeline replaces `make_groups = vars(...)`:

```r
# original DeclareDesign
diagnose_design(design, make_groups = vars(N))

# DeclareDesignZero
simulate_design(design, sims = 500) |>
  group_by(N) |>
  diagnose_design()
```

This generalizes cleanly to any derived column — you can `mutate(sig = p.value < 0.05)` before `group_by(sig)`.

## Variance decomposition

When steps have `draws > 1`, the `$variance_decomposition` slot of the diagnosis object breaks the total variance of per-simulation quantities into one component per draw level plus a residual within-cell component. Labels come from the step names automatically — there are no hardcoded "world" or "design" categories.

```{r vardecomp}
#| cache: true
suppressPackageStartupMessages({
  library(DeclareDesignZero)
  library(fabricatrZero)
  library(randomizr)
  library(estimatr)
  library(dplyr)
})

set.seed(123)
icc <- 0.2

design_vd <- declare_model(
  villages = add_level(N = 20,
               u = draw_normal_icc(clusters = villages, ICC = icc)),
  citizens = nest_level(N = 10,
               Y_Z_0 = u + rnorm(N),
               Y_Z_1 = 0.3 + u + rnorm(N)),
  draws = 100   # 100 distinct worlds (model draws)
) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = villages), draws = 30) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, .method = lm_robust,
                    clusters = villages, inquiry = "ATE", label = "ols")

# 100 × 30 = 3,000 simulation paths; draws= on the steps define the plan
diag_vd <- diagnose_design(design_vd, bootstrap_sims = 0)
```

The decomposition covers `estimate`, `p.value`, `std.error`, and `statistic` by default. For the clustered design above, power, bias, and RMSE can be derived:

```{r vardecomp-table}
#| echo: false
vd <- diag_vd$variance_decomposition

# Derive diagnosand-level quantities from the simulations
sims_vd <- get_simulations(diag_vd) |>
  dplyr::filter(!is.na(estimand)) |>
  dplyr::mutate(
    bias    = estimate - estimand,
    sq_err  = (estimate - estimand)^2,
    sig     = as.numeric(p.value < 0.05)
  )

# Re-run decomposition on the derived quantities
draw_cols   <- grep("_draw$", names(sims_vd), value = TRUE)
group_cols  <- intersect(c("inquiry","estimator","term","outcome"), names(sims_vd))
vd_extended <- DeclareDesignZero:::compute_variance_decomposition(
  sims_vd, draw_cols, group_cols,
  target_cols = c("estimate", "bias", "sq_err", "sig")
)

tbl <- vd_extended |>
  dplyr::select(quantity, var_total, var_model, var_assignment,
                frac_model, frac_assignment) |>
  dplyr::mutate(
    quantity = dplyr::case_when(
      quantity == "estimate" ~ "Estimate",
      quantity == "bias"     ~ "Bias (estimate − estimand)",
      quantity == "sq_err"   ~ "Squared error",
      quantity == "sig"      ~ "Power (p < 0.05)"
    ),
    dplyr::across(dplyr::starts_with("var_"), ~ round(.x, 4)),
    dplyr::across(dplyr::starts_with("frac_"), ~ scales::percent(.x, 0.1))
  )

knitr::kable(tbl, col.names = c("Quantity", "Var(total)", "Var(model)",
                                "Var(assignment)", "% model", "% assignment"),
             align = c("l","r","r","r","r","r"),
             caption = paste0(
               "Variance decomposition: 100 model draws × 30 assignment draws ",
               "= 3,000 simulation paths. draw_levels = model > assignment."
             )) |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = FALSE)
```

The `frac_model` column answers: "what fraction of this quantity's variability comes from which world (population) we're in, versus which randomization we drew?" For RMSE, most variability comes from the model draw — which world we're analyzing matters more than which assignment was realized. For power, the balance is reversed: the randomization draw determines more of the variability than the underlying world.

## Breaking changes

The following changes are the only ones that break code from the DeclareDesign book or the fabricatr documentation. Each is intentional.

### fabricatrZero

#### `add_level(nest = FALSE)` → `declare_level()`

**Old syntax:**
```r
fabricate(
  countries = add_level(N = 20, gdp = runif(N, 1, 10)),
  years     = add_level(N = 10, shock = runif(N, 0, 1), nest = FALSE),
  obs       = cross_levels(by = join_using(countries, years))
)
```

**New syntax:**
```r
fabricate(
  countries = declare_level(N = 20, gdp = runif(N, 1, 10)),
  years     = declare_level(N = 10, shock = runif(N, 0, 1)),
  obs       = cross_levels(.by = c("countries", "years"))
)
```

**Reasoning.** The `nest` argument on `add_level` is a mode switch that makes the same function do two completely different things. When you write two `add_level` calls in sequence, nesting is the only semantically natural outcome — and that is indeed what `add_level(nest = TRUE)` (the default) does. The `nest = FALSE` case is not adding a level to a hierarchy at all: it is declaring an independent population for later crossing. Naming it `declare_level` makes this intent explicit and mirrors the `declare_*` idiom already in the DeclareDesign ecosystem.

The `join_using()` function in the original fabricatr is a fabricatr-internal convenience that mimics dplyr syntax without being dplyr. The `.by = c("names")` replacement is plain character specification, and the implementation uses `dplyr::cross_join()` (added in dplyr 1.1.0) — the correct primitive for Cartesian products.

#### `link_levels(by = join_using(..., rho = r))` → `link_levels(.by = c(...), rho = r)`

**Old syntax:**
```r
students = link_levels(N = 500,
  by = join_using(primary_school, secondary_school, rho = 0.5))
```

**New syntax:**
```r
students = link_levels(N = 500,
  .by = c("primary_school", "secondary_school"), rho = 0.5)
```

**Reasoning.** Embedding `rho` inside `join_using()` conflates two separate concerns: which levels to link (structural) and how correlated the assignments should be (statistical). Separating them makes both independently readable and eliminates the need for a bespoke `join_using` function. The leading dot on `.by` visually distinguishes the control argument from column-generating expressions in the `...`.

#### `modify_level(by = "col")` → `modify_level(.by = "col")`

**Old syntax:**
```r
modify_level(Y_centered = Y - mean(Y), by = "cluster")
```

**New syntax:**
```r
modify_level(Y_centered = Y - mean(Y), .by = "cluster")
```

**Reasoning.** Minor. The leading dot on `.by` is consistent with the `.by` argument convention in dplyr (added in dplyr 1.1.0) and distinguishes grouping control from column expressions. The change is mechanical.

#### `N` is not in the output data frame

In fabricatr, `N` appears as a column in the output data frame when `fabricate(N = k, ...)` is called. In fabricatrZero, `N` is available in the data mask during column evaluation but is not written to the output.

```r
# fabricatr
fabricate(N = 10, Y = rnorm(N))
# → data frame with columns: ID, N, Y

# fabricatrZero
fabricatrZero::fabricate(N = 10, Y = rnorm(N))
# → data frame with columns: Y
```

**Reasoning.** An N column in the output is redundant — it encodes `nrow(df)` as a repeated scalar. It causes concrete problems in cross-classification: when `cross_join` merges two level data frames that both have an N column, dplyr renames them `N...1` and `N...4`, breaking downstream column references. Removing it from the output is cleaner.

### DeclareDesignZero

#### `inquiry =` in `declare_estimator` must be a string, not a step object

**Old syntax (sometimes used):**
```r
my_inquiry <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))
declare_estimator(Y ~ Z, inquiry = my_inquiry)
```

**New syntax:**
```r
declare_estimator(Y ~ Z, inquiry = "ATE")
```

**Reasoning.** The step object `my_inquiry` has an attribute `label = "inquiry"` (the step type), not `"ATE"` (the inquiry name). Passing the step object would silently link the estimator to a non-existent inquiry named `"inquiry"`, producing NA estimand columns with no error. Requiring a string makes the intent explicit and errors immediately if the inquiry name doesn't match.

#### `make_groups = vars(...)` is removed; use `group_by() |> diagnose_design()`

**Old syntax:**
```r
diagnose_design(design, make_groups = vars(N))
```

**New syntax:**
```r
simulate_design(design, sims = 500) |>
  group_by(N) |>
  diagnose_design()
```

**Reasoning.** `make_groups` required understanding a non-standard evaluation convention that differed from standard tidyverse grouping. The pipeline form is more composable: you can mutate derived grouping columns before `group_by`, combine simulations from multiple runs, or filter before diagnosis — none of which `make_groups` supported.

#### `insert_step`, `delete_step`, `replace_step` are deprecated (still work, with a warning)

The step-modification functions are retained for backward compatibility but emit a once-per-session warning pointing toward explicit reconstruction. The recommended replacement is simply to re-declare the design with the desired steps:

```r
# instead of insert_step / replace_step, just rebuild:
design_v2 <- design[1:2] + new_step + design[4:5]
```

#### `declare_potential_outcomes` and `declare_reveal` are not implemented

These functions were deprecated in the original DeclareDesign before this rewrite. The replacement is fabricatrZero's `potential_outcomes()` inside `declare_model`, which achieves the same result with cleaner syntax:

```r
declare_model(
  N = 100,
  potential_outcomes(Y ~ 0.3 * Z + rnorm(N))
)
```

## Code sizes

```{r loc-table}
#| echo: false

loc <- tibble::tibble(
  Package = c("fabricatr", "fabricatrZero", "DeclareDesign", "DeclareDesignZero"),
  `Lines of R` = c(3200, 600, 5400, 1868),
  Files = c(16, 8, 30, 17),
  `Tests` = c("~60", "69", "~100", "187")
)

knitr::kable(loc, align = c("l","r","r","r"),
             caption = "Approximate lines of R code (excluding documentation and tests).") |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover"),
                            full_width = FALSE)
```

The rewrites are roughly 1/3 the size of their originals in both cases, cover the same public API, pass all book examples, and add new capabilities (step-level draws, variance decomposition, `group_by` pipeline, `declare_level`, nested ICC draws in fabricatrZero).
